DCMIP2016, Part 1: Models and Equation Sets

### 2017-04-14 DCMIP2016-Part1-Models.tex
- Cleanup of section 4 - Incorporated changes from FVM and GEM in section 5

This section provide a brief overview of key discretization choices, along with unique features or design specifications from participating dynamical cores.  Further details on these choices can be found in subsequent sections.  Simulation results and model descriptions have been submitted from ten dynamical cores (see Table \ref{tab:Models}).  Although all models solve the Euler equations (or a close approximation of these equations), essentially every model employs a different formulation with a distinct set of prognostic variables.  {\color{red}[Maybe move to a separate paper?] The derivation of many of these formulations can be found in Appendix \ref{sec:EquationSets}.}  The prognostic variables employed and horizontal discretizations for these dynamical cores are summarized in Table \ref{tab:ModelsHorizontal}.  The vertical coordinate chosen by these models and vertical staggering of variables is summarized in Table \ref{tab:ModelsVertical}.  Standard options for diffusion, stabilization, filters, or fixers along with the temporal discretization for these models is summarized in Table \ref{tab:ModelsTemporalDiffusion}.

@@

\begin{table}[h]
\caption{Details on the prognostic variables and horizontal discretization for participating dynamical cores.  Equation set indicates whether a model is hydrostatic (H) or nonhydrostatic (NH), and whether the model presently supports the deep-atmosphere formulation (D).  Only two numerical methods are represented among participating models, namely finite-volume (FV) and spectral-element (SE).  More details on horizontal staggering can be found in section \ref{sec:HorizontalStaggering}.} \label{tab:ModelsHorizontal}
\begin{center}
\begin{tabular}{cccccc}
\hline Acronym & Equation set & Prognostic Variables & Horizontal Grid & Numerical & Horizontal  \\
& & & & method & staggering \\ \hline
CSU & NH (Unified) & $\zeta$, $D$, $w$, $p_s$, $\theta_v$, $q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & Z-grid \\
DYNAMICO & H/NH {\color{red}(D)?} & $\vb{v}_h$, $\rho_s w$, $\rho_s$, $\rho_s \theta_v$, $\Phi$, $\rho_s q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & C-grid \\
FV$^3$ & NH & $\vb{u}_h$, $w$, $\rho_s$, $\rho_s \theta_v$, $\Phi$, $\rho_s q$ & Cubed-sphere (\S \ref{sec:grid_cs}) & FV & D-grid \\
FVM & NH (D) & $\rho_d$, $\rho_d \vb{u}_h$, $\rho_d w$, $\rho_d \theta$, ($\Phi$), $\rho_d q$ & Octahedral reduced (\S \ref{sec:grid_oct}) & FV & A-grid \\
GEM & NH & $\vb{u}_h$, $w$, {\color{red}$\log \partial s / \partial \zeta$}, $\log T_v$, $\Phi$, $q$ & Yin-yang (\S \ref{sec:grid_yinyang}) & FV & C-grid  \\
HOMME & H & $\vb{u}_h$, $\rho_s$, $T_v$, $\rho_s q$ & Cubed-sphere (\S \ref{sec:grid_cs}) & SE & A-grid \\
ICON & NH & $\vb{u}_h$, $w$, $\rho$, $\theta_v$, $\rho q$ & Icosahedral triangular (\S \ref{sec:grid_ico_tri}) & FV & C-grid \\
MPAS & NH & $\rho_d \vb{u}_h$, $\rho_d w$, $\rho_d$, $\rho_d \theta_v$, $\rho_d q$ & CCVT (\S \ref{sec:grid_ccvt}) & FV & C-grid \\
NICAM & NH & $\rho \vb{u}_h$, $\rho w$, $\rho$, $\rho e$, $\rho q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & A-grid \\
Tempest & NH & $\vb{u}_h$, $w$, $\rho$, $\rho \theta_v$, $\rho q$ & Cubed-sphere (\S \ref{sec:grid_cs}) & SE & A-grid \\
\hline 
\end{tabular}
\end{center}
\end{table}

\begin{table}[h]
\caption{Vertical coordinates and staggering for participating dynamical cores.  More details on vertical staggerings can be found in section \ref{sec:VerticalStaggering}.} \label{tab:ModelsVertical}
\begin{center}
\begin{tabular}{cccccc}
\hline Acronym & Vertical Coordinate & Vertical Staggering \\ \hline 
CSU & fixed height & Lorenz \\
DYNAMICO & {\color{red}floating mass?} & Lorenz \\
FV$^3$ & floating mass & Lorenz \\
FVM & (floating) height & Co-located \\
GEM & log pressure & Modified Charney-Phillips (\S \ref{sec:VerticalStaggering})  \\
HOMME & floating mass & Co-located \\
ICON & fixed height & Lorenz \\
MPAS & fixed height & Lorenz \\
NICAM & fixed height & Lorenz \\
Tempest & fixed height & Lorenz \\
\hline 
\end{tabular}
\end{center}
\end{table}

@@

@@

\subsection{NICAM}

@@

In this section we discuss the placement of prognostic variables on the grid and the choice of vertical coordinate.

@@

%----------------------------------------------------
\begin{figure}[]
\includegraphics[width=5in, clip, trim=0cm 2.0cm 0cm 0cm]{HorizontalStaggering.pdf}
\caption{Various options for horizontal staggering of variables among DCMIP models, in this case depicted on a rectilinear grid and geodesic grid.}
\label{fig:horizontalstaggering}
\end{figure}
%----------------------------------------------------

@@

\subsection{Vertical coordinates}

We now present several examples of vertical coordinates that could be used in a non-hydrostatic atmospheric model.

\subsubsection{Terrain-following $z$ coordinates}

In general, terrain-following coordinates are adopted so that the topographic surface is also a coordinate surface.  This simplifies the treatment of the bottom boundary condition.  Perhaps the most common terrain-following coordinate is from \cite{galchen1975}, which takes the form
\begin{align}
s = z_{\footnotesize \mbox{top}} \left( \frac{z - z_s}{z_{\footnotesize \mbox{top}} - z_s} \right).
\end{align}  Analogous formulations are available for mass-based ($\sigma$-coordinates) and entropy-based vertical coordinates.  Because the sharp variations in the coordinate surfaces are preserved far above a rough lower-boundary, new coordinate formulations have been proposed that smooth coordinate surfaces, such as \cite{schar2002new} or \cite{klemp2011terrain}.

@@

Mass-based coordinates are a generalization of pressure-based coordinates to non-hydrostatic models, with a vertical coordinate defined as the total gravity-weighted overhead mass,
\begin{align}
s = \int^{\infty}_{z} \rho g dz.
\end{align}  Under this definition,
\begin{align}
\pdiff{s}{z} = - \rho g.
\end{align}

%The mass-based coordinate gives rise to a prognostic equation for hydrostatic surface pressure through vertical integration of (\ref{eq:PrognosticPseudoDensity}),
%\begin{align}
%\pdiff{}{t} \int \rho_s ds + \int \nabla_s \cdot (\rho_s \vb{u}_h) ds = 0,
%\end{align} where the integrals are taken over the column.

@@

The vertical coordinate of GEM model, denoted $\zeta$, is a hybrid terrain-following coordinate of a log-hydrostatic-pressure type.
Taking $s$ (denoted $\pi$ in GEM documentation) as given in (\ref{eq:s_mass-based}), then $\zeta$ is given by the relation
\begin{align}\label{eq:gem_zeta}
\log s = A(\zeta) + B(\zeta) \left[ \log s(z_s)  - \zeta_s \right],
\end{align}
\noindent with $A(\zeta) = \zeta, \, B(\zeta) =\left( \frac {\zeta-\zeta_{top}} {\zeta_s -\zeta_{top}} \right)^r$
and where  $\zeta_s = \log(10^5)$, $\zeta_{top} = \log(s_{top})$
and $r$ is a variable exponent providing added freedom for adjusting the thickness of model layers over high terrain.

\subsubsection{Floating Lagrangian coordinates} \label{sec:FloatingLagrangian}

@@

\subsection{Model Implementations}

In this section we describe the prognostic equations that have been chosen for the dynamical cores that participated in DCMIP.

@@

DYNAMICO is a shallow-atmosphere fully-compressible model based on a mass-based vertical coordinate and a Hamiltonian formulation (Dubos and Tort, 2014). Prognostic variables are pseudo-density $\rho_s$, mass-weighted tracers (potential temperature, water species), geopotential $\Phi$, ``horizontal'' covariant components of momentum and mass-weighted vertical momentum $W = \rho_s g^{-2} d\Phi/dt = \rho_s g^{-1} w$. Prognostic equations are in flux-form for mass, tracers and $W$, in advective form for $\Phi$ and in vector-invariant form for ``horizontal'' momentum.

@@

{\color{red}[Can this text be more concisely integrated with the text in sections 5.1 - 5.11?]}

The fully compressible Euler equations solved in FVM are given as
\begin{subequations} \label{fvm:compressible}
\begin{align}
 & \pdiff{\mathcal{G}\rho_d}{t} + \nabla \cdot \left(\mathbf{v} \mathcal{G} \rho_d\right) = 0~,
\label{fvm:mass}  \\
& \pdiff{\mathcal{G} \rho_d \mathbf{u}}{t} + \nabla \cdot \left(\mathbf{v} \mathcal{G} \rho_d \mathbf{u} \right) = 
\mathcal{G}\rho_d\left( - \Theta \mathbf{\widetilde{G}} \nabla \phi' - \frac{ \mathbf{g}}{\theta_{a}} \left(\theta'+\theta_a (\varepsilon q'_v - q_c - q_r) \right)
- \mathbf{f} \times \left( \mathbf{u} - \frac{\theta}{\theta_a} \mathbf{u}_{a} \right) 
+ \mathbf{M} \right)~,
\label{fvm:momentum} \\
%- \frac{L}{c_p \pi} & \Big( \frac{\Delta q_{vs}}{\Delta t} & + E_r  \Big)
& \pdiff{\mathcal{G}\rho_d \theta'}{t} + \nabla \cdot \left(\mathbf{v}\mathcal{G}\rho_d\,\theta' \right) = 
\mathcal{G}\rho_d\left(-\mathbf{\widetilde{G}}^{T}\mathbf{u} \cdot \nabla \theta_{a}
- \frac{L}{c_p \pi} \left( \frac{\Delta q_{vs}}{\Delta t}  + E_r  \right) %\left(C_d + E_p\right) 
+  \mathcal{H} \right)~,
\label{fvm:thermodynamic} \\
& \phi' = c_{p} \theta_{0} \left[\left( \frac{R_{d}}{p_{0}} \rho_d \theta (1+q_v/\varepsilon) \right)^{R_{d}/c_{v}} - \pi_{a} \right]~,
\label{fvm:gaslaw}
\end{align}
\end{subequations}
which describe the conservation laws of dry mass \eqref{fvm:mass}, momentum \eqref{fvm:momentum},
and dry entropy \eqref{fvm:thermodynamic}.
Dependent variables in \eqref{fvm:compressible} are dry density $\rho_d$, three-dimensional 
physical velocity vector $\mathbf{u}$, potential temperature perturbation $\theta'$, and Exner 
pressure perturbation $\pi'$, 
with the thermodynamic variables related by the gas law 
\eqref{fvm:gaslaw}~\footnote{Note that $\phi'$ represents a normalised Exner pressure perturbation.}.
Mixing ratios of water vapor, cloud condensate, and rain are denoted as $q_v$, $q_c$, and $q_r$,
respectively. All primed variables correspond to deviations from an ambient state (denoted by
subscript "a")  that satisfies 
a balanced subset of \eqref{fvm:compressible}, thus $\psi'=\psi-\psi_a$, where $\psi = u,v,w,\theta,..$~;
see \cite{prusa2008eulag} and \cite{smolarkiewiczJCP2014}. 
The subscript "0" appearing with $\theta_0$ refers to a constant reference value. 
Symbols appearing on the rhs of the momentum equation \eqref{fvm:momentum}  are the
coefficient 
\begin{equation}
\Theta := \frac{\theta\,(1+q_v/\varepsilon)}{\theta_0\,(1+q_t)} %\equiv \frac{\theta_d}{\theta_0}~,
\end{equation}
in front of the pressure gradient term with $\Theta/\theta_0$ the density potential temperature,
the gravity vector $\mathbf{g} \equiv (0,0,-g)$~\footnote{In the shallow- versus deep-atmosphere 
form of the governing equations,  gravity is constant $g \equiv g_c$ or varies with height as 
$g = g_c\,(a/r)^2$, respectively.}, the Coriolis parameter $\mathbf{f}$, and  
$\varepsilon=1/\varepsilon-1$ with $\varepsilon = R_d/R_v$. % and $\Upsilon_{C} := {\theta}/{\theta_a}$.
The governing equations \eqref{fvm:compressible} are formulated with respect to a geospherical 
coordinate system and a generalised height-based terrain-following vertical coordinate~\footnote{For
simplicity, the vertical coordinate is assumed to be time-independent in the current presentation.}.
Associated symbols are the Jacobian of the metric tensor $\mathcal{G}$, a matrix of metric coefficients
$\widetilde{\mathbf{G}}$, its transpose $\widetilde{\mathbf{G}}^{T}$, 
and the transformation of the physical to the contravariant velocity 
$\mathbf{v} = \mathbf{\widetilde{G}}^{T}\mathbf{u}$; see \cite{prusa2003all} and \cite{kuehnleinJCP2012} for discussion.
%(see \eqref{prusa2003,kuehnleinJCP2012} for discussion). 
The symbol $\mathbf{M}$ in \eqref{fvm:momentum} subsumes metric forces due to the curvature
of the sphere \citep[][]{smolarkiewiczetalJCP2016} and momentum dissipation, whereas $\mathcal{H}$ 
in \eqref{fvm:thermodynamic} represents the diffusion of heat.

%$\mathbf{M}$ and $\mathcal{H}$ subsume various rhs forcings to the
%momentum \eqref{fvm:momentum} and thermodynamic \eqref{fvm:thermodynamic} equations.

@@

GEM is a semi-Lagrangian model with a hybrid coordinate of a log-hydrostatic-pressure type, here denoted $\zeta$,
described in section \ref{sec:GEM_zeta}.
The non-hydrostatic equations are written explicitly as deviations from hydrostatic balance represented by
\begin{align}
\mu = \pdiff{p}{s} - 1,
\end{align}
\noindent where $s$ (denoted $\pi$ in GEM documentation) is given by (\ref{eq:s_mass-based})
and in which case
the equations of GEM model \citep{Girard2014} are:
\begin{align}
\frac{d \vb{u}_h }{dt} + f {\bf k} \times {\bf{u}_{h}} + R_d T_v\nabla_\zeta \,  \log\, p+ \left( 1 + \mu \right) \nabla_\zeta {\Phi} &= 0, \\
\frac{d w}{dt}- g \mu &= 0, \\
\frac{d}{dt} \log \left( \pdiff{s}{\zeta}\right) + \nabla_{ \zeta }\cdot {\bf u}_{{h}}+\frac{\partial \dot{ \zeta }}{\partial \zeta } &= 0, \\
\frac{d \log T_v}{dt} - \frac{R_d}{c_p} \frac {d \log p} {dt} &= 0, \\
\frac{\partial \Phi}{\partial s} + \frac{R_dT_v}{p} &= 0, \\
\frac{d \Phi}{dt}-  g w & = 0.
\end{align}

@@

HOMME uses a floating Lagrange formulation (section \ref{sec:FloatingLagrangian}) for tracer advection and optionally for the vertical dynamics. The 2D vector invariant form of the prognostic horizontal velocity equations (\ref{eq:PrognosticUhEquationK2Sb}) is employed, in conjunction with prognostic temperature (\ref{eq:VirtualTemperaturePrognosticEq}) and surface pressure {\color{red}[Equation?]}.

In addition to the primitive equation model, HOMME also hosts an experimental non-hydrostatic dynamical core which retains many of the features of the traditional CAM-SE model. This model makes use of the shallow-atmosphere Euler equations, as first described by \cite{laprise1992euler}. In its current incarnation, unstructured spectral elements are retained in the horizontal with matching spectral elements in the vertical. Prognostic variables include the 3d wind velocity components $u,v,w$ {\color{red}[Equation?]}, the surface pressure $p_s$ {\color{red}[Equation?]}, potential temperature (\ref{eq:PotentialTemperaturePrognostic}), and geopotential (\ref{eq:ShallowAtmosphereGeopotential}). Fields are co-located in the vertical and explicit time-stepping is used.

@@

%In the horizontal, Tempest utilizes a spectral-element formulation.

@@

Most dynamical cores implement specialized techniques for diffusion or stabilization.  Diffusion is a numerical technique that removes spurious numerical noise from the simulation, where the numerical noise typically arises because of inaccuracies in the treatment of waves with wavelengths near the grid scale.  Stabilization is a numerical technique that prevents energy growth and allows the model to be run over long periods.  Diffusion or stabilization options include physically-motivated turbulence parameterizations, added viscosity or hyperviscosity terms with tunable coefficients, off-centering, or wave-mode filters.  Since the discretization can also lead to an unphysical loss of mass or energy, mass or energy fixers are also employed to replace lost mass or energy to the system.  A comprehensive overview of schemes for diffusion and stabilization schemes can be found in \cite{jablonowski2011pros}.  In this section we discuss some of the diffusion and stabilization strategies employed by the DCMIP suite of dynamical cores.

>>>>

This section provide a brief overview of key discretization choices, along with unique features or design specifications from participating dynamical cores.  Further details on these choices can be found in subsequent sections.  Simulation results and model descriptions have been submitted from ten dynamical cores (see Table \ref{tab:Models}).  Although all models solve the Euler equations (or a close approximation of these equations), essentially every model employs a different formulation with a distinct set of prognostic variables.  The derivation of many of these formulations can be found in Appendix \ref{sec:EquationSets}.  The prognostic variables employed and horizontal discretizations for these dynamical cores are summarized in Table \ref{tab:ModelsHorizontal}.  The vertical coordinate chosen by these models and vertical staggering of variables is summarized in Table \ref{tab:ModelsVertical}.  Standard options for diffusion, stabilization, filters, or fixers along with the temporal discretization for these models is summarized in Table \ref{tab:ModelsTemporalDiffusion}.

@@

\begin{table}[h]
\caption{Details on the prognostic variables and horizontal discretization for participating dynamical cores.  Equation set indicates whether a model is hydrostatic (H) or nonhydrostatic (NH), and whether the model presently supports the deep-atmosphere formulation (D).  Only two numerical methods are represented among participating models, namely finite-volume (FV) and spectral-element (SE).  More details on horizontal staggering can be found in section \ref{sec:HorizontalStaggering}.} \label{tab:ModelsHorizontal}
\begin{center}
\begin{tabular}{cccccc}
\hline Acronym & Equation set & Prognostic Variables & Horizontal Grid & Numerical & Horizontal  \\
& & & & method & staggering \\ \hline
CSU & NH (Unified) & $\zeta$, $D$, $w$, $p_s$, $\theta_v$, $q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & Z-grid \\
DYNAMICO & H/NH {\color{red}(D)?} & $\vb{v}_h$, $\rho_s w$, $\rho_s$, $\rho_s \theta_v$, $\Phi$, $\rho_s q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & C-grid \\
FV$^3$ & NH & $\vb{u}_h$, $w$, $\rho_s$, $\rho_s \theta_v$, $\Phi$, $\rho_s q$ & Cubed-sphere (\S \ref{sec:grid_cs}) & FV & D-grid \\
FVM & NH (D) & $\rho_d$, $\rho_d \vb{u}_h$, $\rho_d w$, $\rho_d \theta$, ($\Phi$), $\rho_d q$ & Octahedral reduced (\S \ref{sec:grid_oct}) & FV & A-grid \\
GEM & NH & $\vb{u}_h$, $w$, {\color{red}$\log \partial s / \partial \zeta$}, $\log T_v$, $q$ & Yin-yang (\S \ref{sec:grid_yinyang}) & FV & C-grid  \\
HOMME & H & $\vb{u}_h$, $\rho_s$, $T_v$, $\rho_s q$ & Cubed-sphere (\S \ref{sec:grid_cs}) & SE & A-grid \\
ICON & NH & $\vb{u}_h$, $w$, $\rho$, $\theta_v$, $\rho q$ & Icosahedral triangular (\S \ref{sec:grid_ico_tri}) & FV & C-grid \\
MPAS & NH & $\rho_d \vb{u}_h$, $\rho_d w$, $\rho_d$, $\rho_d \theta_v$, $\rho_d q$ & CCVT (\S \ref{sec:grid_ccvt}) & FV & C-grid \\
NICAM & NH & $\rho \vb{u}_h$, $\rho w$, $\rho$, $\rho e$, $\rho q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & A-grid \\
Tempest & NH & $\vb{u}_h$, $w$, $\rho$, $\rho \theta_v$, $\rho q$ & Cubed-sphere (\S \ref{sec:grid_cs}) & SE & A-grid \\
\hline 
\end{tabular}
\end{center}
\end{table}

\begin{table}[h]
\caption{Vertical coordinates and staggering for participating dynamical cores.  More details on vertical coordinates can be found in section \ref{sec:VerticalCoordinates}, and details on vertical staggerings can be found in section \ref{sec:VerticalStaggering}.} \label{tab:ModelsVertical}
\begin{center}
\begin{tabular}{cccccc}
\hline Acronym & Vertical Staggering & Vertical Coordinate \\ \hline 
CSU & Lorenz & fixed height \\
DYNAMICO & Lorenz & {\color{red}floating mass?} \\
FV$^3$ & Lorenz & floating mass \\
FVM & Co-located & (floating) height \\
GEM & Modified Charney-Phillips (\S \ref{sec:VerticalStaggering}) & log pressure (\S \ref{sec:GEM_zeta})  \\
HOMME & Co-located & floating mass \\
ICON & Lorenz & fixed height \\
MPAS & Lorenz & fixed height \\
NICAM & Lorenz & fixed height \\
Tempest & Lorenz & fixed height \\
\hline 
\end{tabular}
\end{center}
\end{table}

@@

\subsection{Icosahedral Non-hydrostatic model (ICON)}

@@

\subsection{Non-hydrostatic ICosahedral Atmospheric Model (NICAM)}

@@

Although the horizontal grid determines the approximate density and distribution of grid points, the placement of prognostic (and diagnostic) variables within each grid volume is still free to be specified.  Hence, in this section we turn our attention to the placement of prognostic variables on the grid and the choice of vertical coordinate.

@@

%----------------------------------------------------
\begin{figure}[]
\includegraphics[width=5in, clip, trim=0cm 2.0cm 0cm 0cm]{HorizontalStaggering.pdf}
\caption{Horizontal staggering options represented among DCMIP models, in this case depicted on a rectilinear grid and geodesic grid.}
\label{fig:horizontalstaggering}
\end{figure}
%----------------------------------------------------

@@

\subsection{Vertical Coordinates} \label{sec:VerticalCoordinates}

Most atmospheric models use dimension splitting in order to separate the evolution equations for momentum into horizontal and vertical directions.  Whereas ``horizontal'' in this context either refers to either contravariant velocities, which are perpendicular to the vertical, or covariant velocities, which are directed along coordinate (e.g. terrain-following) surfaces, the vertical is strictly aligned with the radial vector pointing from the center of the Earth.  In this context, the use of dimension splitting allows us to choose an arbitrary vertical coordinate $s$ with the only requirement being that it is a monotonic function of the altitude $z$.  Typically $s$ is chosen so that the Earth's surface (the bottom boundary of the atmosphere) is a coordinate surface, allowing for easy specification of boundary conditions for the prognostic equations -- this leads to the so-called ``terrain-following'' family of vertical coordinates.  Perhaps the most common terrain-following coordinate is from \cite{galchen1975}, which is in terms of the altitude $z$ and takes the form
\begin{align}
s(\vb{x},z) = z_{\footnotesize \mbox{top}} \left[ \frac{z - z_s(\vb{x})}{z_{\footnotesize \mbox{top}} - z_s(\vb{x})} \right],
\end{align} where $\vb{x}$ denotes the horizontal position, $z_s(\vb{x})$ is the height of the topography at that position, and $z_{\footnotesize \mbox{top}}$ denotes the height of the model top (typically independent of position).  Analogous formulations are available for mass-based ($\sigma$-coordinates) and entropy-based vertical coordinates.  Because the sharp variations in the coordinate surfaces are preserved far above a rough lower-boundary, new coordinate formulations have been proposed that smooth coordinate surfaces, such as \cite{schar2002new} or \cite{klemp2011terrain}.  All models in this paper use some variant of terrain-following coordinates, although work on cut-cell, embedded boundary and immersed boundary representations is ongoing (e.g. \cite{lock2012demonstration}).  Note that time-dependent vertical coordinates are allowed and are typically referred to as ``floating'' coordinates.  Several examples of vertical coordinates are now given.

@@

Mass-based coordinates are a generalization of pressure-based coordinates to non-hydrostatic models, with a vertical coordinate defined as the total gravity-weighted overhead mass,
\begin{align} \label{eq:s_mass-based}
s = \int^{\infty}_{z} \rho g dz.
\end{align}  Under this definition,
\begin{align}
\pdiff{s}{z} = - \rho g.
\end{align}

The mass-based coordinate gives rise to a prognostic equation for hydrostatic surface pressure through vertical integration of (\ref{eq:PrognosticPseudoDensity}),
\begin{align} \label{eq:SurfacePressureEquation}
\pdiff{}{t} \int \rho_s ds + \int \nabla_s \cdot (\rho_s \vb{u}_h) ds = 0,
\end{align} where the integrals are taken over the column.

@@

The vertical coordinate of GEM model, denoted $\zeta$, is a hybrid terrain-following coordinate of a log-hydrostatic-pressure type.  Taking $s$ (denoted $\pi$ in GEM documentation) as given in (\ref{eq:s_mass-based}), then $\zeta$ is given by the relation
\begin{align}\label{eq:gem_zeta}
\log s = A(\zeta) + B(\zeta) \left[ \log s(z_s)  - \zeta_s \right],
\end{align}
\noindent with $A(\zeta) = \zeta, \, B(\zeta) =\left( \frac {\zeta-\zeta_{top}} {\zeta_s -\zeta_{top}} \right)^r$
and where  $\zeta_s = \log(10^5)$, $\zeta_{top} = \log(s_{top})$,
and $r$ is a variable exponent providing added freedom for adjusting the thickness of model layers over high terrain.

\subsubsection{Floating Lagrangian coordinates (DYNAMICO, FV$^3$ and HOMME)} \label{sec:FloatingLagrangian}

@@

\subsection{Prognostic equations} \label{sec:PrognosticEquations}

The Navi\'er-Stokes equations that govern atmospheric motion can take on many forms, depending on the choice of prognostic variables and coordinate system.  A derivation of many forms of these equations can be found in Appendix \ref{sec:EquationSets}.  The particular equations used by the model, and the discrete form of these equations, can impact many important aspects of the model, including the presence of computational modes, the accuracy of the model in representing the physical modes of the atmosphere \citep{thuburn2005vertical}, and the ability of the model to conserve important invariants such as energy \citep{Dubos2014Equations}.  The remainder of this section gives some specific examples of prognostic equations used by the DCMIP models.

@@

DYNAMICO is a shallow-atmosphere fully-compressible model based on a mass-based vertical coordinate and a Hamiltonian formulation \citep{Dubos2014Equations}. Prognostic variables are pseudo-density $\rho_s$, mass-weighted tracers (potential temperature, water species), geopotential $\Phi$, ``horizontal'' covariant components of momentum and mass-weighted vertical momentum $W = \rho_s g^{-2} d\Phi/dt = \rho_s g^{-1} w$. Prognostic equations are in flux-form for mass, tracers and $W$, in advective form for $\Phi$ and in vector-invariant form for ``horizontal'' momentum.

@@

FVM solves conservation laws for dry mass \eqref{fvm:mass}, momentum \eqref{fvm:momentum}, and dry 
entropy \eqref{fvm:thermodynamic} in Eulerian flux-form,
which are similar to \eqref{eq:ContinuityEquation} for $\rho_d$, \eqref{eq:PrognosticVelocityFluxForm}, and 
\eqref{eq:PotentialTemperatureFluxForm} for $\theta$, respectively. Underlying the governing equations 
in FVM is a perturbational form with respect to a balanced ambient state, and a generalized curvilinear 
coordinate formulation in a geospherical framework. These governing equations can 
concisely be written as
\citep{smolarkiewiczetalJCP2017}
\begin{subequations} \label{fvm:compressible}
\begin{align}
 & \pdiff{\mathcal{G}\rho_d}{t} + \overline{\nabla} \cdot \left(\mathbf{v} \mathcal{G} \rho_d\right) = 0~,
\label{fvm:mass}  \\
& \pdiff{\mathcal{G} \rho_d \mathbf{u}}{t} + \overline{\nabla} \cdot \left(\mathbf{v} \mathcal{G} \rho_d \mathbf{u} \right) = 
\mathcal{G}\rho_d\left( - \theta_{\rho} \mathbf{\widetilde{G}} \overline{\nabla} \phi' 
- \mathbf{k} \frac{g}{\theta_{a}} \left(\theta'+\theta_a (\frac{q'_v}{\varepsilon} - q'_v - q_c - q_r) \right)
- 2\mathbf{\Omega} \times \left( \mathbf{u} - \frac{\theta_{\rho}}{\theta_{\rho\,a}} \mathbf{u}_{a} \right) 
+ \mathcal{M}' \right)~,
%+ \mathbf{M} \right)~,
\label{fvm:momentum} \\
%- \frac{L}{c_p \pi} & \Big( \frac{\Delta q_{vs}}{\Delta t} & + E_r  \Big)
& \pdiff{\mathcal{G}\rho_d \theta'}{t} + \overline{\nabla} \cdot \left(\mathbf{v}\mathcal{G}\rho_d\,\theta' \right) = 
- \mathcal{G}\rho_d\,\mathbf{\widetilde{G}}^{T}\mathbf{u} \cdot \overline{\nabla} \theta_{a} ~,
%- \frac{L}{c_{pd}\,\pi} \left( \frac{\Delta q_{vs}}{\Delta t}  + E_r  \right) %\left(C_d + E_p\right) 
%+  \mathcal{H} \right)~,
\label{fvm:thermodynamic} \\
& \phi' = c_{pd} \left[\left( \frac{R_{d}}{p_{0}} \rho_d\,\theta\,(1+q_v/\varepsilon) \right)^{R_{d}/c_{vd}} - \pi_{a} \right]~.
\label{fvm:gaslaw}
\end{align}
\end{subequations}
Dependent variables in \eqref{fvm:compressible} are dry density $\rho_d$, 3D
physical velocity vector $\mathbf{u}$, potential temperature perturbation $\theta'$, and a modified
Exner pressure perturbation $\phi'$, 
with the thermodynamic variables related by the gas law 
\eqref{fvm:gaslaw}. Primes indicate perturbations with respect to the prescribed ambient state denoted by
subscript "a"; see also \cite{prusa2008eulag,smolarkiewiczJCP2014} for discussion. 
In the shallow- versus deep-atmosphere form of \eqref{fvm:compressible}, gravity is constant 
$g \equiv g_c$ or varies with height as $g = g_c\,(a/(a+z))^2$, respectively.
As far as geometric aspects are concerned, the nabla operator $\overline{\nabla}$ represents
the 3D vector of partial derivatives with respect to the curvilinear coordinates, along with the Jacobian $\mathcal{G}$, 
a matrix of metric coefficients $\mathbf{\widetilde{G}}$, its transpose $\widetilde{\mathbf{G}}^{T}$, and 
the contravariant velocity $\mathbf{v} = \mathbf{\widetilde{G}}^{T}\mathbf{u}$; see also \cite{prusa2003all} 
and \cite{kuehnleinJCP2012}.
The symbol $\mathcal{M}'(\mathbf{u}, \mathbf{u}_a, \theta_{\rho}/\theta_{\rho\,a})$ in \eqref{fvm:momentum} 
subsumes spherical Christoffel symbols \citep[][]{smolarkiewiczetalJCP2017}.

@@

%GEM is a semi-Lagrangian model with a hybrid coordinate of a log-hydrostatic-pressure type, here denoted $\zeta$, described in section \ref{sec:GEM_zeta}.
In GEM the non-hydrostatic equations are written explicitly as deviations from hydrostatic balance represented by
\begin{align}
\mu = \pdiff{p}{s} - 1,
\end{align}
\noindent where $s$ (denoted $\pi$ in GEM documentation) is given by (\ref{eq:s_mass-based})
and in which case
the equations of GEM model \citep{Girard2014} are:
\begin{align}
\frac{d \vb{u}_h }{dt} + f {\bf k} \times {\bf{u}_{h}} + R_d T_v\nabla_\zeta \,  \log\, p+ \left( 1 + \mu \right) \nabla_\zeta {\Phi} &= 0, \\
\frac{d w}{dt}- g \mu &= 0, \\
\frac{d}{dt} \log \left( \pdiff{s}{\zeta}\right) + \nabla_{ \zeta }\cdot {\bf u}_{{h}}+\frac{\partial \dot{ \zeta }}{\partial \zeta } &= 0, \\
\frac{d \log T_v}{dt} - \frac{R_d}{c_p} \frac {d \log p} {dt} &= 0, \\
\frac{\partial \Phi}{\partial s} + \frac{R_dT_v}{p} &= 0, \\
\frac{d \Phi}{dt}-  g w & = 0.
\end{align}

%\noindent with horizontal velocity ${\bf V}_h$, 
%generalized vertical velocity $\dot{ \zeta }=\frac{d\zeta}{dt}$, 
%vertical velocity $w$,
%geopotential $\phi$, 
%virtual temperature $T_v$, 
%pressure $p$ and
%hydrostatic pressure $\pi$. The variable 
%$\mu$ is a measure of departure from hydrostatic balance, 
%$s$ is related to the surface pressure $\pi_s$ and $B$ is a metric term. The last two variables are defined in (\ref{eq:gem_zeta}).

@@

HOMME uses a floating Lagrange formulation (section \ref{sec:FloatingLagrangian}) for tracer advection and optionally for the vertical dynamics. The 2D vector invariant form of the prognostic horizontal velocity equations (\ref{eq:PrognosticUhEquationK2Sb}) is employed, in conjunction with prognostic temperature (\ref{eq:VirtualTemperaturePrognosticEq}) and surface pressure (\ref{eq:SurfacePressureEquation}).

In addition to the primitive equation model, HOMME also hosts an experimental non-hydrostatic dynamical core which retains many of the features of the traditional CAM-SE model. This model makes use of the shallow-atmosphere Euler equations, as first described by \cite{laprise1992euler}. In its current incarnation, unstructured spectral elements are retained in the horizontal with matching spectral elements in the vertical. Prognostic variables include the 3d wind velocity components $u,v,w$ {\color{red}[Equation?]}, the surface pressure $p_s$ (\ref{eq:SurfacePressureEquation}), potential temperature (\ref{eq:PotentialTemperaturePrognostic}), and geopotential (\ref{eq:ShallowAtmosphereGeopotential}).

@@

@@

Most dynamical cores implement specialized techniques for diffusion or stabilization (see Table \ref{tab:ModelsTemporalDiffusion}).  Diffusion is a numerical technique that removes spurious numerical noise from the simulation, where the numerical noise typically arises because of inaccuracies in the treatment of waves with wavelengths near the grid scale.  Stabilization is a numerical technique that prevents energy growth and allows the model to be run over long periods.  Diffusion or stabilization options include physically-motivated turbulence parameterizations, added viscosity or hyperviscosity terms with tunable coefficients, off-centering, or wave-mode filters.  Since the discretization can also lead to an unphysical loss of mass or energy, mass or energy fixers are also employed to replace lost mass or energy to the system.  A comprehensive overview of schemes for diffusion and stabilization schemes can be found in \cite{jablonowski2011pros}.  In this section we discuss some of the diffusion and stabilization strategies employed by the DCMIP suite of dynamical cores.
